Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-01-10 paper/main.tex
"supplemental" --> "supplementary"
fixed typos identified by R2 (rev 2)

Finally, after watching \textit{Birth of Stars}, predicted knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher than for held-out
incorrectly answered questions. This held when we included all questions in the
analysis ($\U = 38279,~p = 0.022$), when we carried out across-lecture
predictions (\textit{Four Fundamental Forces}: $\U = 6684.5,~p = 0.032$;
\textit{Birth of Stars}: $\U = 6414.5,~p = 0.002$), and and when we carried out
within-lecture knowledge predictions for held-out \textit{Birth of Stars}
questions using other \textit{Birth of Stars} questions from the same quiz and
participant ($\U = 6126,~p = 0.006$). However, we found the \textit{opposite}
effect when we carried out within-lecture knowledge predictions for held-out
\textit{Four Fundamental Forces} questions using other \textit{Four Fundamental
Forces} questions from the same quiz and participant ($\U = 6734,~p = 0.027$).
Specifically, on Quiz 3, our knowledge predictions for held-out correctly
answered questions about \textit{Four Fundamental Forces} were reliably
\textit{lower} than those for their incorrectly answered counterparts.
Speculatively, we suggest that this may reflect participants forgetting some of
the \textit{Four Fundamental Forces} content. If this forgetting happens in a
relatively ``random'' way (with respect to spatial distance within the text
embedding space), then it could explain why some held-out questions about
\textit{Four Fundamental Forces} were answered incorrectly, even if questions
at nearby coordinates (i.e., about similar content) were answered correctly.
This might lead our approach to over-estimate knowledge for held-out questions
about ``forgotten'' knowledge that participants answered incorrectly. Taken
together, the results in Figure~\ref{fig:predictions} indicate that our
approach can reliably predict acquired knowledge (especially about recently
learned content), and that the knowledge predictions are generalizable across
the content areas spanned by the two lectures, while also specific enough to
distinguish between questions about more subtly different content within the
same lecture.

@@

We view our work as making several contributions to the study of how people
acquire conceptual knowledge. First, from a methodological standpoint, our
modeling framework provides a systematic means of mapping out and
characterizing knowledge in maps that have infinite (arbitrarily many) numbers
of coordinates, and of ``filling out'' those maps using relatively small
numbers of multiple choice quiz questions. Our experimental finding that we can
use these maps to predict responses to held-out questions has several
psychological implications as well. For example, concepts that are assigned to
nearby coordinates by the text embedding model also appear to be ``known to a
similar extent'' (as reflected by participants' responses to held-out
questions; Fig.~\ref{fig:predictions}). This suggests that participants also
\textit{conceptualize} similarly the content reflected by nearby embedding
coordinates. The ``spatial smoothness'' of participants' knowledge (as
estimated using quiz performance) is being captured by the knowledge maps we
are inferring from their quiz responses (e.g.,
Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In other words, our
study shows that knowledge about a given concept implies knowledge about
related concepts, and we also show how estimated knowledge falls off with
distance in text embedding space.

@@

At the opposite end of the spectrum from large language models, one could also
imagine \textit{simplifying} some aspects of our LDA-based approach by
computing simple word overlap metrics. For example, the Jaccard similarity
between text $A$ and $B$ is computed as the number of unique words in the
intersection of words from $A$ and $B$ divided by the number of unique words in
the union of words from $A$ and $B$. In a supplemental analysis (Supp.
Fig.~\jaccard), we compared the LDA-based question-lecture matches we reported
in Figure~\ref{fig:question-correlations} with the Jaccard similarities between
each question and each sliding window of text from the corresponding lecture.
As shown in Supplementary Figure~\jaccard, this simple word-matching approach
does not appear to capture the same level of specificity as the LDA-based
approach. Whereas the LDA-based approach often yields a clear peak in the
time series of correlations between each question and the corresponding lecture,
the Jaccard similarity-based approach does not. Furthermore, these LDA-based
matches appear to capture conceptual overlaps between the questions and
lectures (Supp.~Tab.~\matchTab), whereas simple word matching does not. For
example, one of the example questions examined in Supplementary
Figure~\jaccard~asks ``Which of the following occurs as a cloud of atoms gets
more dense?'' The LDA-based matches identify lecture timepoints where the
relevant \textit{topics} are discussed (e.g., when words like ``cloud,''
``atom,'' ``dense,'' etc., are mentioned \textit{together}). The Jaccard
similarity-based matches, on the other hand, are strong when \textit{any} of
these words are mentioned, even if they do not occur together.

@@

For each of 10,000 iterations, we drew a random subsample (with replacement) of 50 participants from our dataset full dataset.
Within each iteration, we first computed the 95\% confidence interval (CI) of the across-subsample-participants mean proportion correct on each of the three quizzes, separately. 
To compute this interval for each quiz, we repeatedly (1,000 times) subsampled participants (with replacement, from the outer subsample for the current iteration) and computed the mean proportion correct each of these inner subsamples. 
We then identified the 2.5\textsuperscript{th} and 97.5\textsuperscript{th} percentiles of the resulting distributions of 1,000 means. 
These three intervals (one for each quiz) served as our thresholds for confidence that the proportion correct within a given distance from a reference question was reliably different (at the $p < 0.05$ significance level) from the average proportion correct across all questions on the given quiz.

>>>>

Finally, after watching \textit{Birth of Stars}, predicted knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher than for held-out
incorrectly answered questions. This held when we included all questions in the
analysis ($\U = 38279,~p = 0.022$), when we carried out across-lecture
predictions (\textit{Four Fundamental Forces}: $\U = 6684.5,~p = 0.032$;
\textit{Birth of Stars}: $\U = 6414.5,~p = 0.002$), and when we carried out
within-lecture knowledge predictions for held-out \textit{Birth of Stars}
questions using other \textit{Birth of Stars} questions from the same quiz and
participant ($\U = 6126,~p = 0.006$). However, we found the \textit{opposite}
effect when we carried out within-lecture knowledge predictions for held-out
\textit{Four Fundamental Forces} questions using other \textit{Four Fundamental
Forces} questions from the same quiz and participant ($\U = 6734,~p = 0.027$).
Specifically, on Quiz 3, our knowledge predictions for held-out correctly
answered questions about \textit{Four Fundamental Forces} were reliably
\textit{lower} than those for their incorrectly answered counterparts.
Speculatively, we suggest that this may reflect participants forgetting some of
the \textit{Four Fundamental Forces} content. If this forgetting happens in a
relatively ``random'' way (with respect to spatial distance within the text
embedding space), then it could explain why some held-out questions about
\textit{Four Fundamental Forces} were answered incorrectly, even if questions
at nearby coordinates (i.e., about similar content) were answered correctly.
This might lead our approach to over-estimate knowledge for held-out questions
about ``forgotten'' knowledge that participants answered incorrectly. Taken
together, the results in Figure~\ref{fig:predictions} indicate that our
approach can reliably predict acquired knowledge (especially about recently
learned content), and that the knowledge predictions are generalizable across
the content areas spanned by the two lectures, while also specific enough to
distinguish between questions about more subtly different content within the
same lecture.

@@

We view our work as making several contributions to the study of how people
acquire conceptual knowledge. First, from a methodological standpoint, our
modeling framework provides a systematic means of mapping out and
characterizing knowledge in maps that have infinite (arbitrarily many) numbers
of coordinates, and of ``filling out'' those maps using relatively small
numbers of multiple choice quiz questions. Our experimental finding that we can
use these maps to predict responses to held-out questions has several
psychological implications as well. For example, concepts that are assigned to
nearby coordinates by the text embedding model also appear to be ``known to a
similar extent'' (as reflected by participants' responses to held-out
questions; Fig.~\ref{fig:predictions}). This suggests that participants also
\textit{conceptualize} similarly the content reflected by nearby embedding
coordinates. How participants' knowledge falls off with spatial distance is
captured by the knowledge maps we infer from their quiz responses
(e.g., Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In other words,
our study shows that knowledge about a given concept implies knowledge about
related concepts, and we also show how estimated knowledge falls off with
distance in text embedding space.

@@

At the opposite end of the spectrum from large language models, one could also
imagine \textit{simplifying} some aspects of our LDA-based approach by
computing simple word overlap metrics. For example, the Jaccard similarity
between text $A$ and $B$ is computed as the number of unique words in the
intersection of words from $A$ and $B$ divided by the number of unique words in
the union of words from $A$ and $B$. In a supplementary analysis (Supp.
Fig.~\jaccard), we compared the LDA-based question-lecture matches we reported
in Figure~\ref{fig:question-correlations} with the Jaccard similarities between
each question and each sliding window of text from the corresponding lecture.
As shown in Supplementary Figure~\jaccard, this simple word-matching approach
does not appear to capture the same level of specificity as the LDA-based
approach. Whereas the LDA-based approach often yields a clear peak in the
time series of correlations between each question and the corresponding lecture,
the Jaccard similarity-based approach does not. Furthermore, these LDA-based
matches appear to capture conceptual overlaps between the questions and
lectures (Supp.~Tab.~\matchTab), whereas simple word matching does not. For
example, one of the example questions examined in Supplementary
Figure~\jaccard~asks ``Which of the following occurs as a cloud of atoms gets
more dense?'' The LDA-based matches identify lecture timepoints where the
relevant \textit{topics} are discussed (e.g., when words like ``cloud,''
``atom,'' ``dense,'' etc., are mentioned \textit{together}). The Jaccard
similarity-based matches, on the other hand, are strong when \textit{any} of
these words are mentioned, even if they do not occur together.

@@

For each of 10,000 iterations, we drew a random subsample (with replacement) of
50 participants from our dataset. Within each iteration, we first computed the
95\% confidence interval (CI) of the across-subsample-participants mean
proportion correct on each of the three quizzes, separately. To compute this
interval for each quiz, we repeatedly (1,000 times) subsampled participants
(with replacement, from the outer subsample for the current iteration) and
computed the mean proportion correct each of these inner subsamples. We then
identified the 2.5\textsuperscript{th} and 97.5\textsuperscript{th} percentiles
of the resulting distributions of 1,000 means. These three intervals (one for
each quiz) served as our thresholds for confidence that the proportion correct
within a given distance from a reference question was reliably different (at
the $p < 0.05$ significance level) from the average proportion correct across
all questions on the given quiz.
